---
title: "Supporting Master’s Students and Supervisors with Metacheck — Project Report"
author:
  - name: "Pr. Dr. Daniël Lakens"
  - name: "Raphael Merz"
  - name: "Dr. Cristian Mesquida"
format: html
toc: true
toc-depth: 3
editor: visual
---

# Motivation and Procedure

The aim of this project was to extend Metacheck, an existing tool for automatically checking research outputs with modules that could be useful to master’s students and their supervisors. Metacheck is a modular, meaning that each module performs a specific type of check. It is available as an [R package](https://www.scienceverse.org/metacheck/) and a [Shiny app](https://shiny.ieis.tue.nl/metacheck/). The motivation was that students often make recurring mistakes in areas such as adherence to reporting guidelines, interpretation of *p*-values or documentation of research practices (e.g., the reporting of a power analysis). Identifying such issues can require substantial time from supervisors, particularly when they need to provide similar feedback repeatedly. Automated checks could therefore help students in identifying and addressing potential issues before submitting their thesis, while simultaneously reducing the burden on supervisors.

By now, Metacheck has developed into an extensible collection of automated checks for research outputs. Inspired by software development, where checks are used to identify potential problems before software is released, Metacheck screens research outputs (e.g., manuscripts and theses) for potential issues and provides information that can help researchers or students follow best practices. Importantly, these checks are intended to support rather than replace human expertise. For example, a check may identify a misreported statistical test or flag a misinterpreted *p*-value, but interpreting the result and deciding whether action is necessary remains the responsibility of the researcher or supervisor.

To better understand which aspects of thesis supervision could benefit from automated checks and how these align with Metacheck’s current modules, we conducted a survey among scientific staff involved in supervising master’s theses in the Department of Industrial Engineering & Innovation Sciences (IE&IS) at TU/e. The results of this survey are presented in the following section.

# Survey of IE&IS Supervisors

The survey was shared in November 2025 with all IE&IS staff members involved in the supervision of master’s theses, with 22 participating. The survey consisted of four open-ended questions addressing (1) the most common types of feedback supervisors provide to master’s students on their theses, (2) which aspects of this feedback they believe could be (partially) automated, (3) which aspects they believe should not be automated, and (4) whether they or their students follow specific reporting guidelines, such as APA. The full survey can be found in Appendix A.

## Survey Results

The survey results indicate that supervisors provide feedback on a broad range of aspects of thesis quality, including both formal presentation and substantive content. Commonly mentioned issues included the coherence and narrative flow of the thesis, the connection between literature and the current study, the formulation and derivation of research questions and hypotheses, and the justification of methodological choices. Supervisors also frequently commented on formatting and structure, the quality of literature summaries and discussions, and citation and reference issues. Statistical reporting was another recurring area of feedback, including checking test assumptions, conducting and reporting power analyses, and providing complete information such as test statistics, degrees of freedom, and exact *p*-values.

Most respondents agreed that at least some of these aspects could be (partially) automated. Formatting, structure and layout were considered particularly suitable, as were checks of research questions, citations, and references. These aspects can generally be assessed against relatively explicit criteria or conventions. However, respondents also identified several areas that they considered less suitable for automation. These mainly involved tasks requiring content expertise, contextual understanding, or critical thinking, such as evaluating methodological choices, developing research questions, and assessing the substantive quality of a thesis. Opinions differed somewhat regarding the overall “storyline” of a thesis: while some considered this inherently dependent on human judgment, others suggested that large language models (LLMs) could potentially help improve the structure and clarity of a thesis. Respondents also emphasized that feedback should be tailored to the individual student and that automated tools should flag potential issues rather than evaluate or grade the overall quality of a thesis. 

Some respondents questioned the desirability of automated feedback, particularly when using LLMs, and raised concerns about privacy and the costs of developing and maintaining such systems. They emphasized that any automated system should therefore be carefully tested before implementation.

Finally, most respondents reported using APA guidelines, while a small number mentioned Harvard or guidelines specific to particular conferences or journals. Overall, the results suggest that automation is most promising for checks based on explicit rules or conventions, whereas tasks requiring contextual judgment and content expertise should remain primarily the responsibility of students and supervisors.

# Using Metacheck

The survey results suggest that that automated checks could complement the existing supervision process. In particular, when supervisors repeatedly provide similar feedback on statistical reporting, citations and references, and the documentation of research practices, automated checks could allow students to detect and address some of these issues independently.

Metacheck is well suited to this role because its checks are designed as supporting tools rather than automated evaluations. The output of a module does not constitute a judgment of the overall quality of a thesis. Instead, it highlights a potential issue, provides guidance, and leaves the final interpretation and decision to the user.

Broadly speaking, Metacheck modules can be classified into four categories: **general modules**, **method modules**, **results modules** and **reference modules**. For a detailed description of all modules, readers are referred to the [Metacheck guide](https://www.scienceverse.org/metacheck_book/). First, the **general modules** check for statements related to Open Science practices, including open and code sharing, conflict of interest, ethics and funding.

Second, the method modules check for preregistrations and the presence and reporting of power analyses.That is, if a power analysis is detected, the corresponding module provides feedback on whether the analysis is fully reported.

The results modules currently provide a broad collection of checks covering adherence to reporting guidelines, including the reporting of and recomputation of effect sizes and *p*-values, interpretation of *p*-values, and best research practices when sharing data and code. Where research outputs are shared in repositories, it can additionally examine the organisation and documentation of data and code, identify potential data-quality and reproducibility problems, and check aspects of compliance with standards for organizing research data.

Finally, the reference modules can identify inconsistencies in reference information and cross-reference cited literature with databases containing information about retractions, replication attempts, and post-publication peer review (i.e., comments on PubPeer). These checks can be useful to identify potentially problematic references, such as papers for which replication attempts exist but are not cited, or papers with PubPeer comments, which may concern methodlogical issues.

The current functionality therefore aligns particularly well with the types of feedback that supervisors identified as potentially automatable: checks based on explicit reporting conventions, common statistical practices, citation and reference information, and other relatively well-defined criteria. At the same time, the survey highlighted several areas that are less suitable for automation, such as evaluating the overall quality of a thesis, judging methodological choices, assessing the substantive contribution of the research, or providing highly context-dependent feedback on its storyline. Metacheck does not attempt to replace this type of supervision. Instead, its role is to identify potential issues that can be checked systematically, allowing students to address more routine problems themselves and enabling supervisors to focus their time on feedback that requires expertise, context, and critical judgment.

Finally, Metacheck can apply multiple modules to a single research output and combine their results into a single report, allowing a thesis and its associated research outputs to be screened across several dimensions at once. This makes it possible to use Metacheck as a first layer of quality control before a supervisor reviews the work in detail. Metacheck can also apply the same set of modules to multiple research outputs simulataneously, allowing supervisors to screen several master's theses in a single analysis.

# Applying Metacheck to Master’s Theses

To illustrate how Metacheck can be applied to master’s theses, we ran the R package `metacheck` on 10 master’s theses manually downloaded as PDFs from the [TU/e library portal](https://research.tue.nl/en/studentTheses/) on 10 August 2026. We searched for master’s theses written in 2025 within the Human-Technology Interaction (HTI) group, which yielded 55 results. We then used the `sample()` function in R to randomly select 10 unique numbers between 1 and 55. The report was written in Quarto using R, including the `metacheck`, `tidyverse` and `kableExtra` packages. All materials, generated data and code required to reproduce this report can be found on Github.

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

# Install the development version of metacheck
pak::pkg_install("scienceverse/metacheck@dev")

# Load packages
library(metacheck)
library(tidyverse)
library(kableExtra)
```

```{r}
#| echo: FALSE
#| eval: FALSE

# Create path to folder with PDF theses
pdf_theses <- file.path("pdf_theses")

# Convert PDF theses to JSON format
json_file <- convert(
  file_path = pdf_theses,
  save_path = "json_theses",
  method = "grobid"
)
```

```{r}
#| echo: FALSE
#| eval: FALSE

# Read in converted JSON file
json_theses <- read(json_file)

# Store for efficiency
saveRDS(json_theses, "results/json_theses.RDS")
```

```{r}
#| echo: FALSE

# Read in converted JSON theses
theses <- readRDS("results/json_theses.RDS")
```

## `ethics_check` module

The `ethics_check` module searches a paper for a statement indicating that the research received **ethics approval**, for example from an Institutional Review Board (IRB), a Research Ethics Committee, or an animal-care committee, or that the research was explicitly exempted from or waived of ethics approval. Reporting ethics approval is generally expected for studies involving human or animal participants.

To illustrate this, we use Thesis 1 and apply the module to identify sentences that may contain an ethics statement. If the module detects an ethics approval statement in a thesis that appears to involve data collection, it returns the following message:

```{r}
#| echo: FALSE

# Run "ethics_check" module
ethics_check <- module_run(theses[1], "ethics_check")
```

> **Ethics Approval Check:** All papers that appeared to involve live data collection contained an ethics approval statement.

@tbl-ethics_check shows the three sentences identified by the module that might refer to an ethics statement. In the first case, the detected sentence is indeed an ethics statement.

```{r}
#| echo: FALSE
#| tbl-cap: "Potential ethics statements detected by the `ethics_check` module"
#| label: tbl-ethics_check

# Display results as a table
ethics_check$table |> 
  filter(section_type == "method" & ethics == TRUE) |> 
  select(
    text
  ) |> 
  kable() |> 
  column_spec(1, width = "15cm") 
```

## `stat_p_nonsig` module

The `stat_p_nonsig` module finds every non-significant *p*-value (*p* $\geq$\$\\geq\$ .05) so a reader can check whether the surrounding text interprets it correctly. A common error is to treat a non-significant result as evidence that there is no effect (“there was no difference between groups, *p* = .27”), when a non-significant result is simply inconclusive.

@tbl-stat_p_nonsig shows the first five non-significant *p*-values detected by the module, along with the sentence in which each *p*-value appears. This allows users to assess whether the *p*-value may have been misinterpreted. In this case, none of the sentences include a statement indicating that there was no difference.

```{r}
#| echo: FALSE
#| tbl-cap: "Non-significant *p*-values detected by the `stat_p_nonsig` module"
#| label: tbl-stat_p_nonsig

# Run "stat_p_nonsig" module
p_nonsig <- module_run(theses, "stat_p_nonsig")

# Display results as a table
p_nonsig$table |> 
  filter(section_type == "results") |> 
  select(
    text, 
    expanded
    ) |> 
  head(5) |> 
  kable() |> 
  column_spec(1, width = "2cm") |> 
  column_spec(2, width = "13cm")
```

## `marginal` module

The `marginal` module lists sentences describing an effect as “marginally significant” (or a related phrase). Describing a non-significant result as “marginally significant”, “approaching significance”, or “trending toward significance” is a common way to misinterpret a non-significant *p*-value. The module detects these sentences so the reader can judge each case in context.

```{r}
#| echo: FALSE

# Run "marginal" module
marginal <- module_run(theses, "marginal")
```

> **Marginal Significance:** You described 5 effect with terms related to 'marginally significant' If *p* values above 0.05 are interpreted as an effect, you inflate the alpha level, and increase the Type 1 error rate. If a *p* value is higher than the prespecified alpha level, it should be interpreted as a non-significant result.

@tbl-marginal shows the *p*-values that were detected as being interpreted as “marginally significant” across the ten theses in the Results section.

```{r}
#| echo: FALSE
#| label: tbl-marginal
#| tbl-cap: "*P*-values detected as being interpreted as marginally significant by the `marginal` module"

# Display results as a table
marginal$table |> 
  filter(section_type == "results") |> 
  select(
    text, 
    section_type
    ) |> 
  kable() |> 
  column_spec(1, width = "12cm") |> 
  column_spec(2, width = "3cm")
```

## `stat_effect_size` module

The `stat_effect_size` module checks whether effect sizes are reported for *t*-tests and *F*-tests. Reporting a standardized effect size alongside a test is recommended by the APA Journal Article Reporting Standards, and makes the result more readily usable in meta-analyses.

Running the `stat_effect_size` module on the ten theses, returns the following warning:

```{r}
#| echo: FALSE

# Run "stat_effect_size" module
effect_size <- module_run(theses, "stat_effect_size")
```

> **Effect Sizes in t-tests and F-tests**: We found 26 t-tests and/or F-tests where effect sizes are not reported. Check these tests in the table below, and consider adding effect sizes

@tbl-stat_effect_size shows the statistical tests detected by the module in Thesis 1, including the extracted test, the effect size and text from which it was extracted. For all five statistical tests, no standardized effect size was reported as indicated by the `NA` values in the `es` column.

```{r}
#| echo: FALSE
#| label: tbl-stat_effect_size
#| tbl-cap: "Statistical tests detected by the `stat_effect_size` module"

# Display results as a table
effect_size$table |> 
  filter(paper_id == 1) |> 
  select(
    test_text, 
    es, 
    text
    ) |>
  kable() |> 
  column_spec(1, width = "3cm") |> 
  column_spec(2, width = "1cm") |> 
  column_spec(3, width = "10cm")
```

The `stat_effect_size` module also checks whether the reported effect size is coherent with the reported test statistic and degrees of freedom. The module recomputes what the effect size should be from the test statistic and degrees of freedom. It checks Cohen’s *d* and *g* for *t*-tests, and $\eta_p^2$ and $\omega_p^2$ for *F*-tests. @tbl-stat_effect_size_coherence shows the first two detected *F*-tests and the recomputed effect size in Thesis 6. If no standardized effect size is detected, then the `es` column appears as `NA`.

```{r}
#| echo: FALSE
#| label: tbl-stat_effect_size_coherence
#| tbl-cap: "Coherence of standardized effect sizes reported along with an *F*-test by the `stat_effect_size` module"

# TODO CM: Can a mismatch occur solely due to rounding of effect size?

# Display results as a table
effect_size$table |> 
  filter(paper_id == 6) |> 
  slice_head(n = 2) |> 
  select(
    test_text, 
    es, 
    eta_implied_partial,
    eta_coherence_note
    ) |> 
  kable() |> 
  column_spec(1, width = "3cm") |> 
  column_spec(2, width = "2cm") |> 
  column_spec(3, width = "4cm") |> 
  column_spec(4, width = "5cm")
```

## `stat_check` module

The `stat_check` module recomputes the *p*-value from each reported test statistic and its degrees of freedom. It flags cases where the reported *p*-value is inconsistent with the test statistic and degrees of freedom and, more seriously, where the inconsistency changes the statistical significance decision (i.e., a decision error, such as claiming significance when the recomputed *p*-value is not significant). The `stat_check` module return results only for *t*-tests and *F*-tests and detects only statistics reported in APA format, which is commonly used in psychology journals. @tbl-stat_check shows the five *p*-values identified as reporting errors by the module after recomputation.

```{r}
#| echo: FALSE
#| label: tbl-stat_check
#| tbl-cap: "Errors in the reporting of *p*-values detected by `stat_check` module"

# Run "stat_check" module
stat_check <- module_run(theses, "stat_check")

# Display results as a table
stat_check$table |> 
  filter(error == "TRUE") |> 
  select(
    test_type, 
    reported_p, 
    computed_p, 
    error, 
    decision_error
) |> 
  kable() |> 
  column_spec(1, width = "2cm") |> 
  column_spec(2, width = "2cm") |> 
  column_spec(3, width = "2cm") |> 
  column_spec(4, width = "2cm")
```

## `repo_check` module

The `repo_check` module identifies links to research repositories in research outputs including, the Open Science Framework (OSF), GitHub, ResearchBox, PsychArchives, and Zenodo. It then retrieves the list of files in each and reports what has been shared. @tbl-repo_check_summary displays the summary of the modules, including the number of repositories, files, data files, code files, readme files and naming issues. We can see that two theses reported links to repositories.

```{r}
#| eval: FALSE
#| echo: FALSE

# Run "repo_check" module
repo_check <- module_run(theses, "repo_check")

# Save for efficiency
saveRDS(repo_check, "results/res_repo_check.RDS")
```

```{r}
#| echo: FALSE
#| label: tbl-repo_check_summary
#| tbl-cap: "Summary of the `repo_check` module"

# Load results
repo_check <- readRDS("results/res_repo_check.RDS")

# Display results as a table
repo_check$summary_table |> 
  select(
    paper_id, 
    repo_n, 
    files_n, 
    files_data, 
    files_code, 
    files_readme, 
    naming_issues
    ) |> 
  kable()
```

The module also detects naming issues in repository files, such as spaces in file names, which can cause problems when sharing code and data. To illustrate some of the issues that can be detected in file naming, we use Thesis 3, for which the module detected naming issues, as shown in @tbl-repo_check_summary. @tbl-repo_check_naming shows the first five naming issues detected by the module in Thesis 3, along with the type of issue and an explanation of each issue.

```{r}
#| echo: FALSE
#| label: tbl-repo_check_naming
#| tbl-cap: "Naming issues detected by the `repo_check` module in Thesis 3"

thesis_four <- module_run(theses[4], "repo_check")

# Display results as a table
thesis_four$naming_issues |> 
  select(
    file_name,
    rule,
    detail
  ) |> 
  head(5) |> 
  kable() |> 
  column_spec(1, width = "6cm") |> 
  column_spec(2, width = "3cm") |> 
  column_spec(3, width = "5cm") 
```

## `power` module

The \`power\` module uses regular expressions to identify sentences describing statistical power analyses and classifies each as a priori, sensitivity, or post-hoc. When collecting data, best practice is to conduct an a priori power analysis to ensure that study design has adequate statistical power to detect the effect size of interest.

@tbl-power shows the three power analyses detected by the module, including the extracted sentence, the section were it was detected and the type of power analysis assigned. Two could not be classified (i.e., were labelled “unknown”) because they did not contain terms such as “a priori”, “sensitivity” or “post-hoc” while one was classified as “a priori”. It is worth noting that the third power analysis is not a power analysis but rather a statement about how small sample sizes may result in inadequate statistical power.

```{r}
#| echo: FALSE
#| label: tbl-power
#| tbl-cap: "Power analyses detected by the `power` module"

# Set to FALSE to opt out from LLM use
llm_use(FALSE)

# Run power module
power <- module_run(theses, "power")

# Display results as a table
power$table |> 
  select(
    text,
    section_type,
    power_type
  ) |> 
  kable() |> 
  column_spec(1, width = "9cm") |> 
  column_spec(2, width = "3cm") |> 
  column_spec(3, width = "2cm")
```

To assess the completeness of power analysis reporting, we rely on an LLM to extract and classify information from reported power analysis statements. This can be enabled by setting `llm_use(TRUE)`. The LLM is also used to classify potential power statements that do not contain terms such as “a priori,” “sensitivity,” or “post-hoc.” When enabled, the module reads each power analysis statement and extracts structured information, including the statistical test, planned sample size, alpha level, desired power, and effect size. It then uses this information to assess whether the power analysis is fully reported.

```{r}
#| eval: FALSE
#| echo: FALSE

# Enables LLM support
llm_use(TRUE)

# Specify model
llm_model("groq/llama-3.3-70b-versatile") 

# Run power module with LLM support
res_llm_power <- module_run(theses, "power")

# Save results for efficiency
saveRDS(res_llm_power, "results/res_llm_power.RDS")
```

```{r}
#| eval: FALSE
#| echo: FALSE
#| label: tbl-power_llm
#| tbl-cap: "Classification and completeness of power analyses by the `power` module using LLM functionality"

# Load results
res_llm_power <- readRDS("results/res_llm_power.RDS")

# Display results as a table
res_llm_power$table
```

## `ref_pubpeer` module

The `ref_pubpeer` module cross-references the cited references against [PubPeer](https://pubpeer.com), a platform for post-publication peer review. The module reports which cited references have PubPeer comments, how many, and a link, so readers can read the discussion and judge its relevance themselves. Comments can range from minor requests for clarification to serious concerns about the research. Therefore, identifying cited articles with potentially serious concerns raised on PubPeer can provide useful information for students and supervisors to consider when evaluating the cited literature. @tbl-ref_pubpeer shows the cited articles with PubPeer comments detected by the module, along with their corresponding DOIs and links to the PubPeer discussions.

```{r}
#| echo: FALSE
#| label: tbl-ref_pubpeer
#| tbl-cap: "PubPeer comments on cited articles detected using the `ref_pubpeer module`"

# Run module
pubpeer <- module_run(theses, "ref_pubpeer")

# Display results as a table
pubpeer$table |> 
  select(
    doi, 
    total_comments, 
    url
    ) |> 
  head(3) |> 
  kable() |> 
  column_spec(1, width = "3cm") |> 
  column_spec(2, width = "3cm") |> 
  column_spec(3, width = "8cm")
```

# Summary

Overall, these examples illustrate how Metacheck can be used to identify potential issues in master’s theses from both the student’s perspective, as a tool for identifying and addressing common issues before submission, and the supervisor’s perspective, as a first layer of quality control that can help identify areas requiring further attention. However, automated tools such as Metacheck are not infallible and may produce both false positives and false negatives. The results should therefore be interpreted as indicators of potential issues rather than definitive judgments. Ultimately, it remains the responsibility of the student or supervisor to review the flagged issues, assess their relevance in context, and determine whether any action is required.

```{r}
#| echo: FALSE

# Create a text file with session information
capture.output(
  sessioninfo::session_info(),
  file = "session-info.txt"
)
```

{{< pagebreak >}}

# Appendix A: Survey 

**Survey on Automated Feedback for Master Theses** 

Dear participant,

\
We are developing new modules for a tool designed to help master students and their supervisors check for common mistakes and improve adherence to best research practices in theses. This tool currently detects issues such as missing sample size justifications, incorrect statistical interpretations of non-significant p-values, and references to retracted papers. 

This survey aims to identify common feedback provided by supervisors on master’s theses, focusing on recurring mistakes related to structural aspects, research methods and statistical reporting, and citations. Additionally, we also seek to understand which aspects of feedback supervisors believe could be automated to better assist students in the process. 

Examples of these automated checks include:

- verifying the presence/absence of section headers;
- analyzing the number of references and their citation frequency;
- and checking whether tables and figures are referenced in the main text.  

To develop these new modules, we are exploring different approaches, including rule-based text searches (e.g., identifying specific phrases indicating common mistakes) and using large language models (LLMs) to detect issues and provide automated suggestions. Note that we only use LLM and API providers with high data security standards, ensuring that any data processed by the tool will not be shared with third parties or used for model training.

Your input will help shape the development of this tool, ensuring that it addresses real challenges faced by students and supervisors. The survey should take approximately  **5 minutes**  to complete. 

Thank you for your time and insights!

Q1: What are some of the most common types of feedback you provide to master students on their theses? Please focus on recurring issues related to the structure of the master thesis (e.g., missing headings), research methods (e.g., not reporting statistical tests, poor reporting practices (not reporting exact *p*-values), citations (missing reference), or other structural aspects.

Q2: Are there any aspects of the feedback process that you think could be (partially) automated?

Q3: Are there any aspects of thesis feedback that you believe should not be automated?

Q4: Do you or your students follow any field-specific reporting guidelines when writing/critiquing the thesis?  For instance, social scientists and psychologists often use the American Psychological Association (APA) guidelines to format manuscripts and report results. 

Q5: Is there anything else you would like to share with us?
